---
title: 'STAT167 HW3 - Spring 2022'
author: "Victoria Nguyen"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---

***

## **Homework #3 instructions**

**Review [textbook Chapter 3 "Data visualization"](https://r4ds.had.co.nz/data-visualisation.html) and lecture notes on `ggplot2` (2022/4/19 & 2022/4/21) before answering the homework questions**.

This homework contains 2 questions, each with multiple parts, 100 points in total.   

<font color="blue">Replace **INSERT_YOUR_ANSWER** with your own answers.</font>

- First open this `rmd` file in RStudio and click `Knit -> Knit to PDF` to render it to PDF format. You need to have `LaTex` installed on the computer to render it to PDF format. If not, you can also render it to HTML format.
- It is best to read this `rmd` file and the rendered `pdf`/`html` file side-by-side, while you are working on this homework.
- If the question asks you to write some R code, remember to put your code into a **R code chunk**. Make sure both your R code chunk and its output are visible in the rendered `html` file.
- For this homework, use **ggplot2** to visualize your data. Do **NOT** use R base graphics functions.

- <font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

- If you have any question about this homework assignment, we encourage you to post it on [**Piazza**](http://piazza.com/ucr/spring2022/stat167/home).

### **Homework submission guideline**

- **This homework is DUE at _11:59 PM_ on _Sunday April 24, 2022_**.
- Late submission penalties.  
    - Submissions up to 24 hours late will incur a 10% deduction.  
    - Submissions up to 48 hours late will incur a 30% deduction.

- <font color="blue">**If you are using one or both of your free late days, please state here: **</font> **INSERT_YOUR_ANSWER**

- After you complete all questions, save your `rmd` file to `FirstnameLastname-SID-HW3.rmd` and save the rendered pdf file to `FirstnameLastname-SID-HW3.pdf`. If you can not knit it to pdf, knit it to html first and then print/save it to pdf format. 
- Submit **BOTH your source `rmd` file and the knitted `pdf` file** to **GradeScope**. Do NOT create a zip file. For the `pdf` submission, please tag specific pages that correspond with each question in the assignment.
- You can submit multiple times, you last submission will be graded.



***

## **Acknowledgments**

<font color="blue">
Please list all the help you have received for completing this homework.
</font>

[](https://ggplot2.tidyverse.org/reference/theme.html)

***

**Load necessary packages**

```{r, collapse=T}
# install the tidyverse package first if you have not done it yet.
#install.packages("tidyverse") # you can comment out this line after you have installed `tidyverse`

library(tidyverse) # for the `ggplot2` package
```


***

## **Question 1 [40pt] Visualization of the `mpg` data set**

This data set contains fuel economy data 1999 - 2008 for 38 popular car models (source: EPA [http://fueleconomy.gov](http://fueleconomy.gov)).

```{r, collapse=T}
?mpg
dim(mpg)  # dimension of the table
names(mpg)  # list the variables in mpg
str(mpg)  # list the structures in mpg
glimpse(mpg) # get a glimpse of the mpg data
```

### (a) [20pt] Visualize the distribution of drive train types

(i) Draw a barplot (frequency histogram) to display the distribution of `drv`, the type of drive train. Use different colors to distinguish different drive train types. Explicitly label the number of car models of each drive train type on top of the bars.


```{r, collapse=T}
ggplot(data = mpg, mapping = aes(x = drv, fill = drv)) + # x axis and color using drv
  geom_bar() + 
  geom_text(stat = 'count', # using count (frequency)
            mapping = aes(label = ..count..), # adds labels for each count 
            vjust = -.5) # moves the label to slightly above the bars 
```

(ii) Draw a coxcomb or pie chart to display the proportions of each drive train types.

```{r, collapse=T}
ggplot(data = mpg) + 
  geom_bar(mapping = aes(x = drv, fill = drv), #same data and coloring as previous plot
           width = 1) + 
  coord_polar() # changes to polar coordinate chart 
```

***

### (b) [20pt] How highway mileage varies across drive train type?

Generate a horizontal boxplot to compare the distribution of highway mileage across three different drive train types. Reorder the boxes by the median mileage values.

```{r, collapse=T}
ggplot(data = mpg) + 
  geom_boxplot(mapping = aes(x = reorder(drv, hwy, FUN = median), y = hwy)) + # generate boxplot where  x = drv, y = hwy, then order using median
  coord_flip() # flip to horizontal
```


***

\newpage

## **Question 2 [60pt] Visualization the `diamonds` data set**

This data set contains the prices and other attributes of almost 54,000 diamonds. 

```{r, collapse=T}
?diamonds
dim(diamonds)  # dimension of the table
diamonds # print/view diamonds
str(diamonds)  # list the structures in diamonds
glimpse(diamonds) # get a glimpse of the data
```

### (a) [20pt] Heatmap of `cut` vs `clarity`

(i) Use [the `geom_tile()` function](https://ggplot2.tidyverse.org/reference/geom_tile.html) to make a heatmap to visualize the number of diamonds in each `cut` and `clarity` combination.

```{r, collapse=T}
ggplot(data = count(x = diamonds, cut, clarity)) + #using cut and clarity data from diamonds
geom_tile(mapping = aes(x = cut, y = clarity, fill = n)) # x = cut, y = clarity, fill using count
```

(ii) Change the color palette of your heatmap.

**Hint**: See [these heatmap examples at the R Graph Gallery](https://www.r-graph-gallery.com/79-levelplot-with-ggplot2.html).


```{r, collapse=T}
ggplot(data = count(x = diamonds, cut, clarity)) + 
geom_tile(mapping = aes(x = cut, y = clarity, fill = n)) + #tile x = cut, y = clarity, fill = count
  scale_fill_gradient(low = "darkgreen", high = "lightyellow") #custom gradient used for fill 
```

***

### (b) [40pt] Visualize the distribution of diamond price

(i) Use [the `geom_histogram()` function](http://ggplot2.tidyverse.org/reference/geom_histogram.html) to compare the distribution of `price` across different `cut`. Change the y-axis to density, and use the `dodge` position adjustment.


```{r, collapse=T}
ggplot(data = diamonds) + 
  geom_histogram(aes(x = price, y = ..density.., fill = cut), position = "dodge")

```

(ii) Use [the `geom_freqpoly()` function](http://ggplot2.tidyverse.org/reference/geom_histogram.html) to compare the distribution of `price` across different `cut`. Change the y-axis to density.



```{r, collapse=T}
ggplot( data = diamonds) + 
  geom_freqpoly(aes(x = price, y = ..density.., color = cut)) # freqpoly x = price, y = density, and color = cut

```

(iii) Use [the `geom_boxplot()` function](https://ggplot2.tidyverse.org/reference/geom_boxplot.html) to compare the distribution of `price` across different `cut`.


```{r, collapse=T}
ggplot(data = diamonds) + 
  geom_boxplot(mapping = aes(x = cut, y = price)) # vertical boxplot x = cut y = price
```

(iv) Use [the `geom_violin()` function](https://ggplot2.tidyverse.org/reference/geom_violin.html
) to compare the distribution of `price` across different `cut`.

```{r, collapse=T}
ggplot( data = diamonds)+ 
  geom_violin(mapping = aes(x = cut, y = price)) # violin x = cut y = price
```

(v) What observations can you make from the above plots? Which visualization function is your favorite? Explain your choice.

The plots show that diamonds can range from low to high price, but the a premium cut is what makes a a diamond's price go up the most. The histogram plot is the most straightforward to me, however I also found the heatmap interesting as a way to visualize many intersections. 