---
title: 'STAT167 Lab #4 - Spring 2022'
author: "Victoria Nguyen"
date: "2022/4/21"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: 3
  word_document:
    toc: yes
    toc_depth: '3'
---

***

# **Discussion week 4 instructions**

This week, we will review some `ggplot2` example figures. 


- First, download the `rmd` file from Canvas. 
- Open this `rmd` file in RStudio and click `Knit -> Knit to PDF` to render it to PDF format. You need to have `LaTex` installed on the computer to render it to PDF format. If not, you can also render it to HTML format.
- Read this `rmd` file and the rendered `pdf`/`html` file side-by-side, to see how this document was generated!
- Be sure to play with this document! Change it. Break it. Fix it. The best way to learn R Markdown (or really almost anything) is to try, fail, then find out what you did wrong.

- Read over the `ggplot2` example code and check the output. If you have any questions about certain functions or parameters, it is the time to ask!
- There are some exercises through out this document. Replace **INSERT_YOUR_ANSWER** with your own answers. Knit the file, and check your results.

<font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

**Lab submission guideline**

- After you completed all exercises, save your file to `FirstnameLastname-SID-lab4.rmd` and save the rendered pdf file to `FirstnameLastname-SID-lab4.pdf`. If you can not knit it to pdf, knit it to html first and then print/save it to pdf format.
- Submit **BOTH your source `rmd` file and the knitted `pdf` file** to **GradeScope**. Do NOT create a zip file. 
- You can submit multiple times, you last submission will be graded.

***

# **Lecture Review - `ggplot2`**

## Install the The `tidyverse` package

```{r, collapse=T}
# install the tidyverse package first if you have not done it  yet.
#install.packages("tidyverse") # you can comment out this line after you have installed `tidyverse`
library(tidyverse)
```

## The `mpg` data set

This data set contains fuel economy data 1999 - 2008 for 38 popular car models (source: EPA. [fueleconomy.com/](http://fueleconomy.gov)).

```{r, collapse=T}
?mpg
dim(mpg)  # dimension of the table
mpg  # print/view mpg (we will introduce tibble later in this class)
str(mpg)  # list the structures in mpg
glimpse(mpg) # get a glimpse of the mpg data
```


## The complete graphing template in `ggplot2`
```{r, eval=F}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>,        # optional
     position = <POSITION> # optional
  ) +
  <COORDINATE_FUNCTION> +  # optional
  <FACET_FUNCTION> +       # optional
  <SCALE_FUNCTION> +       # optional
  <THEME_FUNCTION>         # optional
```

\newpage

## Example: `geom_boxplot()` for one variable

Let's first take a look at the distribution of highway mileage.

```{r, fig.height=5.5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = "", y = hwy))+
  facet_wrap(~class)
```

### **Exercise #1**

If you add `facet_wrap(~class)` to the above code, what will happen?

It creates a box plot for each class and displays them in a grid

\newpage

## Example: `geom_boxplot()` for two variables

How highway mileage varies across car classes?

- Plot the distribution of highway mileage by class.

```{r, fig.height=5.5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy))  
```

To make the trend easier to see, we can reorder `class` based on the median value of `hwy`.

```{r, fig.height=5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), 
                             y = hwy))
```

When you have long variable names, `geom_boxplot()` will work better if you flip it with `coord_flip()`.

```{r, fig.height=5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = reorder(class, hwy, FUN = median), 
                             y = hwy)) +
  xlab("class") +
  coord_flip()
```  

### **Exercise #2**

Can you draw a violin plot to compare the distribution of highway mileage varies across car classes? Order your violins by the median values of highway mileage and flip the plot to horizontal. 

```{r}
viol_1<- ggplot(mpg, aes(x= reorder(class, hwy, FUN = median), 
                         y=hwy)) + 
  geom_violin()

viol_1 + coord_flip()

```

### **Exercise #3**

There are two ways to zoom in a ggplot. Compare the following two plots. What are the differences? Which zooming option do you prefer? Explain your reasons.

In the first plot, we can still see all of the boxplots, though not in their entirety (ex: pickup, suv). These are both cropped out in the 2nd lot as the plot is zoomed too far in. I like the first plot more as it's more informative and too much is left out in the 2nd plot. 

```{r, collapse=T, fig.height=5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy)) +
  ylim(c(20, 30))
```  

```{r, collapse=T, fig.height=5}
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy)) +
  coord_cartesian(ylim=c(20, 30))
```  

\newpage

## Example: `geom_histogram()` to compare highway mileage across car classes

```{r, collapse=T, fig.height=5}
ggplot(data = mpg) +
  geom_histogram(mapping = aes(x = hwy, fill = class),
                 binwidth = 10,
                 position = "dodge")
```

### **Exercise #4**

(a) Why there are gaps in the histogram?

The gaps are where there were no cars with those highway miles per gallon (30, 40 mpg). It also seems like there is a border around the histogram (ie, the gap between 0 on the y-axis and the x-axis, as well as large margins on the sides, especially the left) because the graph isn't scaled to fill/make the most of the space in the box. 

The gap in the bottom might be for a "rug" though (ex jittered rug plot)

(b) Can you change this frequency histogram to a density histogram?

```{r}

ggplot(data = mpg) +
  geom_histogram(mapping = aes(x = hwy, y = ..density.., fill = class),
                 binwidth = 10,
                 position = "dodge")
```
\newpage

## Example: `geom_freqpoly()` to compare highway mileage across car classes

```{r, collapse=T, fig.height=5}
ggplot(data = mpg) +
  geom_freqpoly(mapping = aes(x = hwy, color = class),
                binwidth = 10)
```