---
title: "STAT167 Lab #3 - Spring 2022"
author: "Victoria Nguyen"
date: "2022/4/14"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: 3
---

***

# **Discussion week 3 instructions**

This week, we will review some `ggplot2` example figures from the lectures. 

- First, download the `rmd` file from Canvas. 
- Open this `rmd` file in RStudio and click `Knit -> Knit to PDF` to render it to PDF format. You need to have `LaTex` installed on the computer to render it to PDF format. If not, you can also render it to HTML format.
- Read this `rmd` file and the rendered `pdf`/`html` file side-by-side, to see how this document was generated!
- Be sure to play with this document! Change it. Break it. Fix it. The best way to learn R Markdown (or really almost anything) is to try, fail, then find out what you did wrong.

- Read over the `ggplot2` example code and check the output. If you have any questions about certain functions or parameters, it is the time to ask!
- There are some exercises through out this document. Replace **INSERT_YOUR_ANSWER** with your own answers. Knit the file, and check your results.

<font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

**Lab submission guideline**

- After you completed all exercises, save your file to `FirstnameLastname-SID-lab3.rmd` and save the rendered pdf file to `FirstnameLastname-SID-lab3.pdf`. If you can not knit it to pdf, knit it to html first and then print/save it to pdf format.
- Submit **BOTH your source `rmd` file and the knitted `pdf` file** to **GradeScope**. Do NOT create a zip file. 
- You can submit multiple times, you last submission will be graded.

***

# **Lecture Review - `ggplot2`**

## Install the `tidyverse` package

```{r, collapse=T}
# install the tidyverse package first if you have not done it  yet.
#install.packages("tidyverse") # you can comment out this line after you have installed `tidyverse`
library(tidyverse)
```

## The `mpg` data set

This data set contains fuel economy data 1999 - 2008 for 38 popular car models (source: EPA. [fueleconomy.com/](http://fueleconomy.gov)).

```{r, collapse=T}
?mpg
dim(mpg)  # dimension of the table
mpg  # print/view mpg (we will introduce tibble later in this class)
str(mpg)  # list the structures in mpg
glimpse(mpg) # get a glimpse of the mpg data
```


## The complete graphing template in `ggplot2`
```{r, eval=F}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>,        # optional
     position = <POSITION> # optional
  ) +
  <COORDINATE_FUNCTION> +  # optional
  <FACET_FUNCTION> +       # optional
  <SCALE_FUNCTION> +       # optional
  <THEME_FUNCTION>         # optional
```

## Example: aesthetic mappings for both `geom_point()` and `geom_smooth()`

```{r, collapse=T}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = hwy, col = drv) # global mapping
       ) +
  geom_point() +
  geom_smooth(mapping = aes(linetype = drv)) # local mapping for the layer
```

### **Exercise #1**
Write your own `ggplot2` code to make a scatterplot of `cty` against `displ`; add a regression line to your plot; map `drv` to the color aesthetic for both points and lines; map `cyl` to the `size` aesthetic for points. 

```{r}
ggplot(data = mpg, 
       mapping = aes(x = displ, y = cty, col = drv) #global mapping
       ) +
  geom_point(mapping = aes(color = drv, size = cyl))+
  geom_smooth()+ # local mapping for the layer\
  geom_smooth(method='lm', formula= y~x, color = "orange")+
  geom_point(aes(size=cyl))

```
What you can conclude from your figure?

As engine displacement increases, the number of cylinders also increases. Cars with very few (4) cylinders tend to have the best city miles per gallon and all have less than 3 litres of displacement.Rear wheel drive tends to have higher displacement and city mpg below 20, 4 wheel drive has displacement ranging from 2.75 to 6.5 litres, and front wheel drive ranges from 1.5 to 5.5 litres of displacement but overall has the best city mpg. 

## Facets - making subplots that each display one subset of the data.

### Facets by a single variable - `facet_wrap()`

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

### Facets by a combination of two variables - `facet_grid()`

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

### **Exercise #2**

(a) Look at the following code and output figure. What happens if you facet on a continuous variable?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cyl, y = hwy)) + 
  facet_wrap(~ displ, nrow=4)
```

It creates a graph for every different value of x (here, cyl)

(b) Look at the following code and output figures. What does `.` do?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ class)
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(class ~ .)
```

changes the orientation of the graphs, (class ~ .) is horizontal (wide), (. ~ class) is vertical (tall). The x and y axes remain in the same orientation, it just changes the width and height of the graphs. 

## The `diamonds` data set

This data set contains the prices and other attributes of almost 54,000 diamonds. 

```{r, collapse=T}
?diamonds
dim(diamonds)  # dimension of the table
diamonds # print/view diamonds
str(diamonds)  # list the structures in diamonds
glimpse(diamonds) # get a glimpse of the data
```

### **Exercise #3** 
Which variables in `diamonds` are categorical? Which variables are numerical? 

Categorical:  Cut, Color, Clarity
Numerical:    Carat, Depth, Table, Price, X, Y, Z

## geom vs stat functions

Statistical transformation is a alternative way to build a layer.

Most geoms and stats come in pairs that are almost always used in concert. 

- every geom has a default stat
- every stat has a default geom. 

You can typically use geoms without worrying about the underlying statistical transformation. 

You can also use geoms and stats interchangeably.

## Example: `geom_bar()` calls `stat_count()` make bar plot for a categorical variable 

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

## Override default stat function

To make a proportion bar chart (relative frequency histogram), we need to override the default `group` argument and map the variable `..prop..` (calculated by `stat_count() `) to the `y` aesthetic.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop.., group = 1))
```


## Example: position adjustment options for `geom_bar()` 

```{r, collapse=T}
?geom_bar
```

The position argument specifies the position adjustment of bars, rectangles.

- **default**: position = "stack"
- position = "identity" will place each object exactly where it falls in the context of the graph. 
- position = "fill" works like stacking, but makes each set of stacked bars the same height. 
- position = "dodge" places overlapping objects directly beside one another. 
the bars are automatically stacked. Each colored rectangle represents a combination of cut and clarity.


```{r, collapse=T, fig.height=5}
#install.packages("gridExtra") # you only need to install it once
library(gridExtra)
gg <- ggplot(data = diamonds,
             mapping = aes(x = cut, fill = clarity)) 
plot1 <- gg + geom_bar(position="stack")
plot2 <- gg + geom_bar(position="identity")
plot3 <- gg + geom_bar(position="fill")
plot4 <- gg + geom_bar(position="dodge")
grid.arrange(plot1, plot2, plot3, plot4, ncol=2)
```
