---
title: "STAT167 HW1 - Spring 2022"
author: "Victoria Nguyen"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---

***

## **Homework #1 instructions**

**Review 2022/3/31, 2022/4/5, and 2022/4/7 lecture notes before answering the homework questions**.

This homework contains 2 questions, each with multiple parts, 100 points in total. 

<font color="blue">Replace **INSERT_YOUR_ANSWER** with your own answers.</font>

- First open this `rmd` file in RStudio and click `Knit -> Knit to PDF` to render it to PDF format. You need to have `LaTex` installed on the computer to render it to PDF format. If not, you can also render it to HTML format.
- It is best to read this `rmd` file and the rendered `pdf`/`html` file side-by-side, while you are working on this homework.
- If the question asks you to write some R code, remember to put your code into a **R code chunk**. Make sure both your R code chunk and its output are visible in the rendered `html` file.
- For this homework, use **R Base Graphics** to generate the figures. Do **NOT** use `ggplot2` for this homework.

- <font color="blue">**Please comment your R code thoroughly, and follow the R coding style guideline (https://google.github.io/styleguide/Rguide.xml). Partial credit will be deducted for insufficient commenting or poor coding styles.**</font>

- If you have any question about this homework assignment, we encourage you to post it on [**Piazza**](http://piazza.com/ucr/spring2022/stat167/home).

### **Homework submission guideline**

- **This homework is DUE at _11:59 PM_ on _Sunday April 10, 2022_**.
- Late submission penalties.  
    - Submissions up to 24 hours late will incur a 10% deduction.  
    - Submissions up to 48 hours late will incur a 30% deduction.

- <font color="blue">**If you are using one or both of your free late days, please state here: **</font> **INSERT_YOUR_ANSWER**

- After you complete all questions, save your `rmd` file to `FirstnameLastname-SID-HW1.rmd` and save the rendered pdf file to `FirstnameLastname-SID-HW1.pdf`. If you can not knit it to pdf, knit it to html first and then print/save it to pdf format. 
- Submit **BOTH your source `rmd` file and the knitted `pdf` file** to **GradeScope**. Do NOT create a zip file. For the `pdf` submission, please tag specific pages that correspond with each question in the assignment.
- You can submit multiple times, you last submission will be graded.

***

## **Acknowledgments**

<font color="blue">
Please list all the help you have received for completing this homework.
</font>

[https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/cor]

[https://statsandr.com/blog/outliers-detection-in-r/]

[http://www.sthda.com/english/wiki/abline-r-function-an-easy-way-to-add-straight-lines-to-a-plot-using-r-software]

[https://www.youtube.com/watch?v=XaNKst8ODEQ]

[https://www.geeksforgeeks.org/change-legend-size-in-base-r-plot/]
***

**Load necessary packages**

```{r, collapse=T}
library(datasets) # for the `quakes` data set

## You need to install the maps package first, then you can comment out the following line.
# install.packages("maps")

library(maps) # for map visualization
```


***

## **Question 1 [70pt] Analysis of the `quakes` data set**

The `quakes` data set give the locations of 1000 seismic events of MB > 4.0. The events occurred in a cube near Fiji since 1964.

\footnotesize
```{r}
library(datasets) # we have already loaded the package but we can do it again
?quakes 
```

```{r, collapse=T}
class(quakes)
head(quakes, n=5) # print first 5 rows of quakes
dim(quakes)  # dimension of the table
names(quakes)  # list the variables in quakes
str(quakes)  # list the structures in quakes
```

### (a) [10pt] Visualizations of the earthquake magnitudes

Suppose we are interested in studying the distribution of the magnitude `mag` variable. Write your own R code to make a single figure with the following four subfigures in a 2-by-2 layout.  

- subfigure #1: plot a density histogram of the earthquake magnitudes, and then plot the estimated probability density curve in red color in the same plot
- subfigure #2: plot a horizontal boxplot of the earthquake magnitudes
- subfigure #3: plot the empirical cdf of the earthquake magnitudes
- subfigure #4: make a Q-Q plot to compare the observed earthquake magnitudes distribution with the theoretical Normal distribution. Add a *thick* qqline in blue color.  

[**Note**: make sure each subfigure has proper axis labels and figure title.]

```{r, collapse=T}
# 4 figures in a 2x2 layout 
par(mfrow=c(2,2))

# creating variable containing only magnitudes
quake_magnitudes <- quakes$mag

#creating the histogram
hist(quake_magnitudes, freq = FALSE, xlab = "Magnitude", main = "Density Histogram: Magnitudes")

#calculating density
mag_density <- density(quake_magnitudes)
#adding density to histogram
lines(mag_density, lwd = 2, col = "red")

#subfigure 2: horizontal boxplot of earthquake magnitudes
boxplot(quakes$mag, xlab = "Magnitude", horizontal=TRUE, main = "Boxplot")

#SUBFIGURE 3: EMPIRICAL CDF OF EARTHQUAKE MAGNITUDES
mag_ecdf <- ecdf(quake_magnitudes)
plot(mag_ecdf, main = "Empirical CDF")

#SUBFIGURE 4: Q-Q PLOT COMPARING OBSERVED MAGNITUDE DISTRIBUTION WITH THEORETICAL NORMAL DISTRIBUTION. 
qqnorm(quakes$mag, pch = 1, frame = FALSE)
#adding thick qqline in blue 
qqline(quakes$mag, col = "blue", lwd = 3)
```

### (b) [10pt] Outlier detection

There are several outliers labeled by open circles in the boxplot. Write a simple code to print out the indexes of the outliers and then print out the outlier observations.  

[**Hint**: you can use the `summary()` function to find out $Q_1$ and $Q_3$.]

summary(mag_density$y)

Outliers are points that fall below `Q1 - 1.5*IQR` or above `Q3 + 1.5*IQR`

```{r}
#retrieving outliers from the boxplot
boxplot.stats(quakes$mag)$out
out <- boxplot.stats(quakes$mag)$out

#retrieving info from original data about the outliers
out_ind <- which(quakes$mag %in% c(out))
out_ind
```
How many outliers have you found?

There are 7 outliers

### (c) [10pt] Minimum variance unbiased estimates for Normal distribution

Suppose we assume that the earthquake magnitude variable follows a Normal distribution and our 1000 earthquake magnitude observations are independent and identically distributed (iid). 

That is, $X_i \sim N(\mu, \sigma^2)$, where $X_i$ is the magnitude of the $i$-th earthquake observation and $i=1,2,\cdots,1000$.  

The **minimum variance unbiased estimators (MVUE)** for $\mu$ and $\sigma^2$ are:
$$
\widehat{\mu}=\overline{X}=\frac{1}{n}\sum_{i=1}^n X_i \\
\widehat{\sigma^2}=S^2=\frac{1}{n-1}\sum_{i=1}^n (X_i - \overline{X})^2
$$

Use existing R functions to calculate the estimates of $\widehat{\mu}$ and $\widehat{\sigma^2}$.

```{r}
var(quakes$mag)
mean(quakes$mag)
```
Do NOT use any existing R functions, write your own R code to calculate the estimates of $\widehat{\mu}$ and $\widehat{\sigma^2}$.
```{r}
mag_sum <- sum(quakes$mag)
n <- length(quakes$mag)

#calculating mean
mu_hat1 <- mag_sum/n

# calculating variance
sum_sq <- sum((quakes$mag - mu_hat1)^2)
var_hat1 <- sum_sq/(n-1)

#printing results
mu_hat1
var_hat1
```


### (d) [10pt] Visualize your model fitness

To visualize the model fitness, you can add the estimated Normal distribution curve to the histogram plot you have generated in part (a).

Write your own code to generate the following figure.  

- make a density histogram first, set `xlim` from 0 to 8.
- plot the empirical density curve in red color on the same figure.
- plot the estimated Normal distribution curve (that is, $X \sim N(\widehat{\mu}, \widehat{\sigma^2})$) as a blue dashed line on the same figure.
```{r}
density_mag <- density(quakes$mag)

#plotting density histogram 
hist(quake_magnitudes, freq = FALSE, xlab = "Magnitude", main = "Density Histogram: Magnitudes", xlim = c(0,8), ylim = c(0,1.2))

#adding empirical density curbe in red
lines(density_mag,lwd = 2, col = "red")

#adding normal distribution curve in blue
curve(dnorm(x, mean = mean(density_mag$x), sd = sd(density_mag$y)), add=TRUE, col ="blue", lty = "dashed")
```


How does your estimated normal curve differ from the empirical density curve? Do you think the earthquake magnitude observations follow a Normal distribution? Does this result consist with the Q-Q plot you generated in part (a).

My estimated normal curve peaks around the same density as the empirical density curve, however the estimated distribution curve is offset to the right (higher magnitudes).I think this is consistent with the Q-Q plot as we can see that more points fell above the line than below it. I do think the earthquake magnitude observations follow a normal distribution. 

### (e) [10pt] Pairwise scatterplots

Run the following code. 
```{r}
# require(graphics) # load the graphics library

pairs(quakes, main = "Fiji Earthquakes, N = 1000", cex.main=1.2, pch=".")
```

Describe the output figure. 

The figure is a matrix of scatterplots. The diagonal shows the labels for the data used, and each of the other cells shows the combination of the intersection. For example, the 2nd cell in the first row shows the correlation between latitude and longitude, the 3rd cell shows the correlation between latitude and depth, etc. 

Do you think the `mag` variable and `stations` variables are positively correlated? Explain your answer.

Yes, I think `mag` and `stations` are positively correlated as the figures at their intersections (row 4, cell 5 and row 5, cell 4) have positive slopes. 


### (f) [10pt] Single scatterplot 

Write your own R code to reproduce the scatterplot for the `mag` and `stations` variables.  

- calculate the covariance and correlation coefficient. [**Hint**: look up the `cor` package.]
- redraw the outlier points you identified in (b) using red color filled diamond symbol.
- add a three-line legend to your plot. The first legend line reports the covariance value; the second legend line reports the correlation coefficient value; and the third legend line indicates the red diamonds are likely outliers.

```{r}
#scatterplot for `mag` and `stations` variables 
plot(quakes$mag, quakes$stations, main="Scatterplot of Magnitude & Stations",
   xlab="Richter Magnitude", ylab="Number of Stations Reporting", pch=1, lwd = 1) 

#calcuating covariance and correlation using the cor package
magstat_cov <- cov(quakes$mag, quakes$stations, use = "everything",
    method = c("pearson", "kendall", "spearman"))
magstat_cov

magstat_cor <- cor(quakes$mag, quakes$stations, use = "everything",
    method = c("pearson", "kendall", "spearman"))
magstat_cor

x <- quakes$mag
y <- quakes$stations
#redrawing outlier points as red-filled diamonds
points(x[x > 5.8], y[x > 5.8], col = "red",pch=18, cex = 1.5)

#adding three-line legend : covariance, correlation coefficient, and one to mark outliers 
abline(v=7.508181, lty =2, col = "blue")
abline(lm(y ~ x), lty =3, col = "blue")
legend("bottomright", legend=c("Likely Outliers", "Covariance = 7.508181", "Correlation Coefficient = 0.8511824"), col = c("red","black","blue"),pch=c(18,3,3), cex = c(.8))
```


### (g) [10pt] Earthquake maps

We can plot our earthquake records on a world map using the `maps` package. Look at the following map. Where is Fiji?

The Fiji Islands are to the East of Australia, where earthquakes are very dense. 

```{r, collapse=T}
library(maps)
map()
points(quakes$long, quakes$lat, pch=".", col="red")

```

We can also zoom in and make local map of Fiji and its neighborhood area.

Add our earthquake locations into the above plot as red color dots. Use blue color filled rectangles to mark locations of the outliers identified in part (b).

```{r, collapse=T}
long.mean <- mean(quakes$long)
lat.mean <- mean(quakes$lat)
#orient <- c(lat.mean, long.mean, 0)
xlim <- c(min(quakes$long)/2, max(quakes$long)*1.5)
ylim <- c(min(quakes$lat)-10, max(quakes$lat)+10)
map(database="world", xlim=xlim, ylim=ylim, col="grey80", fill=T)

#adding earthquakes as red dots
points(quakes$long, quakes$lat, pch=".", col="red")

#parsing outliers from earthquake data
quakes_outliers <- quakes[out_ind, ]
quakes_outliers

#marking them as blue rectangles
points(quakes_outliers$long, quakes_outliers$lat, pch=15,cex = .8, col="blue")
```

***

## **Question 2 [30pt] Introductory Survey**

Take a few minutes to complete the following short survey. _For multiple choice questions, mark appropriate answers using **bold** text_. Note that as long as your complete this survey, you will receive full credit for this question.

### (a) What is your major and concentration (for your degree)?

I am majoring in Statistics with a concentration in Statistical Computation
 
### (b) What is your minor (if any)?

Public Policy
 
### (c) Why did you decide to take STAT167?

It is part of the Computation concentration, but I am also pursuing work in Data Science after graduation.

### (d) What would you like to get out of this course?

I am hoping this class will help me identify the skills needed for work in data science while also exposing me to said skills. I have done some data science projects outside of school but I feel like I don't have a good idea of the general tools needed in the industry.

### (e) What operating system(s) do you run on your primary computer(s)?  
[Mark your answer in **bold**].  

**Mac OS X**

### (f) What programming languages do you know?   
[Mark all that apply in **bold**].  

- BASIC
- C
**- C++**
- C#
**- Java**
- JavaScript
- HTML / CSS
- LISP (lisp, scheme, clojure, etc)
- Perl
- PHP
**- Python**
- Ruby
**- SQL**
**- VB / VBScript**
**- R**
**- SAS**
**- Matlab**
- Julia
- Scala
- Other: please specify

### (g) What is your primary programming language?  

C++, Python, R

### (h) How often do you write code (before taking this class)?  
[Mark your answer in **bold**].  

**- daily**
- two or more time per week
- once per week
- less than once per week
- little to no programming experience
 
### (i) Overall, how comfortable are you with programming?  
[Mark your answer in **bold**].  

- 1 (less comfortable)
- 2
- 3
**- 4**
- 5 (very comfortable)

### (j) What probability and statistics courses have you already taken, if any?

- Elements of Probability & Statistical Theory
- General Statistical Models
- Regression Analysis
- Statistical Computing with SAS
- Introduction to Statistical Computing
- Introduction to Statistics

### (k) What computer science courses have you already taken, if any?
- Introduction to Data Structures and Algorithms
- Introduction to Computer Science for Science, Mathematics, and Engineering II
